Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-08-05 pot/LaTeX/jsm2014.tex
mapping2
updates
generate sample datasets and plot them
bugfix predictions
Found a fairly major bug in the mcmc

\setbeamercolor{structure}{fg=wp-red}
\setbeamercolor{title}{bg=white, fg=wp-red}  % changes color on title page
\setbeamerfont{title}{series=\bfseries, size=\huge}
\setbeamerfont{author}{series=\bfseries, size=\Large}
\setbeamerfont{institute}{series=\mdseries, size=\large}

@@

\title[Spatiotemporal Modeling of Extreme Events] % (optional, use only with long paper titles)
{
  Spatiotemporal Modeling of Extreme Events
}
\author[S. Morris and B. Reich]{Samuel Morris and Brian Reich}
\institute[NCSU]{North Carolina State University}
\date{}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
    \end{itemize}
    \item In geostatistical analysis, kriging uses spatial correlation to help inform prediction at unknown locations.
    \item Want to explore computationally easy methods that are available in higher dimensions
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard non-spatial analysis}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Block maxima:
    \begin{itemize}
      \item Uses yearly maxima
      \item Discards many observations
      \item Models are fit using the generalized extreme value distribution
    \end{itemize}
    \item Generalized extreme value distribution (GEV):
    \begin{align*}
      \Pr(Y_j < y) = G_j(y) = \exp \left\{ -\left[ \left(1 + \xi_j \frac{y - \mu_j}{\sigma_j}\right)_+^{-1/\xi_j} \right] \right\}
    \end{align*}
    \item For a spatial analysis, max-stable processes give an appropriate limiting distribution.
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis}
  \begin{itemize}
    \item Peaks-over-threshold:
    \begin{itemize}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and fit data above the threshold using the generalized Pareto distribution
      \item Autocorrelation may be an issue between observations (e.g. flood levels don't dissipate overnight)
    \end{itemize}
    \item Generalized Pareto distribution (GPD):
    \begin{align*}
      \Pr(Y_j > y | Y_j > T) = F_j(y) = \left( 1 + \xi_j \frac{y - T}{\sigma_j} \right)_+^{-1/\xi_j}
    \end{align*}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Multivariate analysis}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Multivariate max-stable and GPD models have nice features, but they are
    \begin{itemize}
      \item computationally hard to work with
      \item joint distribution only available in low dimension
    \end{itemize}
    \item Pairwise likelihood approach (Huser and Davison, 2014)
  \end{itemize}
\end{frame}

\begin{frame}{Model objectives}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Our objective is to build a model that
    \begin{itemize}
      \item has a flexible tail
      \item has asymptotic spatial dependence
      \item computation on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{$\chi$ coefficient}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item The $\chi$ coefficient is a measure of extremal correlation
   \item Specifically, we focus on $\chi(c, \bh)$ for the upper tail given by
    \begin{align*}
      \chi(c, \bh) = \Pr(Y(\bs) > c \mid Y(\bs + \bh) > c)
    \end{align*}
    \item If $\lim_{c \rightarrow \infty} \chi(c) = 0$, then high extreme values are not correlated.
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistical models, observations, $Y(\bs)$ are often modeled using a Gaussian spatial model with a likelihood of
      \begin{align*}
    	\hspace{-2em}
    	f(y(\bs)) = \frac{ 1 }{ (2 \pi)^{n/2} } | \Sigma |^{-1/2} \exp \left[ -\frac{ 1 }{ 2 } (Y(\bs) - \mu(\bs) )^T \Sigma^{-1} (Y(\bs) - \mu(\bs)) \right]
    \end{align*}
    where
    \begin{itemize}
    	\item $\mu(\bs)$ is a mean vector
	\item $\Sigma$ is a covariance matrix
    \end{itemize}
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(c) = 0$ regardless of the strength of the correlation in the bulk of the distribution.
    \item Tail is not flexible
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item \alert{Conditioned} on $z_t$ and $\sigma^2_t$, $Y_t(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model.
    \item Predictions can be made through kriging.
    \item \alert{Marginalizing} over $z_t$ and $\sigma^2_t$ (via MCMC),
    \begin{align*}
      Y_t(\bs) \sim \text{skew-t}(\mu, \Sigma^*, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
    	\item $\mu$ is the location
	\item $a$, $b$ are the IG parameters for $\sigma^2_t$
	\item $\Sigma^* = \frac{ b }{ a } \Sigma$ is a scale matrix, and $\Sigma$ is a \Matern covariance matrix
	\item $\alpha \in \calR$ controls the skewness
    \end{itemize}

  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by $\alpha$ and degrees of freedom
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(c) > 0$ Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(c, \bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$ 
    \item Deal with this through a daily random partition.
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $\{w_{t1}, \ldots, w_{tK}\}$ that define a daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - w_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MVT distribution as before.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Example daily partition}
	Two sample partitions
    \centering
    \includegraphics[width=0.54\linewidth]{./plots/example-partition-1.pdf}
    \includegraphics[width=0.54\linewidth]{./plots/example-partition-2.pdf}
\end{frame}

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute missing observations and data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods incorporating
    \begin{itemize}
      \item Gaussian vs $t$ vs skew-$t$ marginal distribution
      \item $K=1$ partition vs $K=5$ partitions
      \item No thresholding vs thresholded
      \begin{itemize}
        \item Thresholded data at $T=0.90$ sample quantile
      \end{itemize}
    \end{itemize}
    \item All methods use a \Matern or exponential covariance ($\nu = 0.5$)
    \item Compare quantile and Brier scores using 5-fold cross validation (Gneiting and Raftery, 2007)
    \item Mean function modeled using a first-order spatial trend
  \end{itemize}
\end{frame}

\begin{frame}{Quantile score}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item The quantile score for the $\tau$th quantile is
    \begin{align*}
      2 \{ I[y < \widehat{q}(\tau)] - \tau\} (\widehat{q} - y)
    \end{align*}
    where:
    \begin{itemize}
      \item $y$ is a test set value
      \item $\widehat{q}(\tau)$ is the estimated $\tau$th quantile
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Brier score}
  \begin{itemize} \setlength{\itemsep}{0.5em}
	\item Brier score for predicting exceedance of threshold $c$
	\begin{align*}
	  [e(c) - P(c)]^2
	\end{align*}
	where 
	\begin{itemize}
		\item $y$ is a test set value
		\item $e(c) = I[y > c]$
		\item $P(c)$ is the predicted probability of exceeding $c$
	\end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Future work}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Comparison with extreme value analysis methods
    \item Reporting of compliance with EPA ozone standards
    \item Spatial mapping of high quantiles
    \item Including time in the model
    \begin{itemize}
      \item AR(1): $Y_t(\bs) = X_t(\bs) \beta + \phi Y_{t-1}(\bs) + \alpha z_t(\bs) + v_t(\bs)$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Questions}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Any questions?
    \item Thank you for your attention.
  \end{itemize}
\end{frame}

>>>>

\setbeamercolor{structure}{fg=wp-red}
\setbeamercolor{title}{bg=white, fg=wp-red}  % changes color on title page
\setbeamerfont{title}{series=\bfseries, size=\huge}
\setbeamerfont{author}{series=\bfseries, size=\large}
\setbeamerfont{institute}{series=\mdseries, size=\large}

@@

\title[Spatiotemporal Modeling of Extreme Events] % (optional, use only with long paper titles)
{
  Spatiotemporal Modeling of Extreme Events
}
\author[S. Morris, B. Reich, and D. Cooley]{Samuel Morris (NC State), Brian Reich (NC State), and Dan Cooley (Colorado State)}
\institute[NCSU, CSU]{}
\date{JSM 2014, Boston}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture.
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels.
    \end{itemize}
    \item Spatial extreme borrows strength across space to estimate return levels and make predictions at unknown locations..
    \item In geostatistical analysis, kriging uses spatial correlation for prediction.
    \item Want to explore similar methods for extremes.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Standard analysis - Block maxima}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Uses yearly maxima
      \item Discards many observations
      \item Models are fit using the generalized extreme value distribution
    \item For a spatial analysis, max-stable processes give an appropriate limiting distribution
  \end{itemize}
\end{frame}

\begin{frame}{Standard analysis - Peaks over threshold}
  \begin{itemize}
      \item Incorporates more data than block maxima
      \item Select a threshold, $T$, and fit data above the threshold using the generalized Pareto distribution
      \item Temporal dependence may be an issue between observations (e.g. flood levels don't dissipate overnight)
    \item Generalized Pareto distribution (GPD) is used for the exceedances.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Multivariate analysis}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Multivariate max-stable and GPD models have nice features, but they are
    \begin{itemize}
      \item computationally challenging to work with
      \item joint distribution only available in low dimension
    \end{itemize}
    \item Pairwise likelihood approach (Huser and Davison, 2014)
  \end{itemize}
\end{frame}

\begin{frame}{Model objectives}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Our objective is to build a model that
    \begin{itemize}
      \item has marginal distribution with a flexible tail
      \item has asymptotic spatial dependence
      \item has computation on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{$\chi$ coefficient}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item The $\chi$ coefficient is a measure of extremal dependence
   \item Specifically, we focus on $\chi(\bh)$ for the upper tail given by
    \begin{align*}
      \chi(\bh) = \lim_{c \rightarrow \infty}\Pr(Y(\bs) > c \mid Y(\bs + \bh) > c)
    \end{align*}
    \item If $ \chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
    \item We expect $\lim_{\bh \rightarrow \infty}\chi(h)=0$.
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function
$\rho(\bh)$.
    \item Model properties:
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
    \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(c) = 0$ regardless of the strength of the correlation in the bulk of the distribution.
    \item Tail is not flexible (Gaussian is light tailed)
    \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item \alert{Conditioned} on $z_t$ and $\sigma^2_t$, $Y_t(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model.
    \item Predictions can be made through Kriging.
    \item \alert{Marginalizing} over $z_t$ and $\sigma^2_t$ (via MCMC),
    \begin{align*}
      Y_t(\bs) \sim \text{skew-t}(\mu, \Sigma^*, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
    	\item $\mu$ is the location
	\item $a$, $b$ are the IG parameters for $\sigma^2_t$
	\item $\Sigma^* = \frac{ b }{ a } \Sigma$ is a scale matrix, and $\Sigma$ is a \Matern covariance matrix
	\item $\alpha \in \calR$ controls the skewness
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail controlled by skewness $\alpha$ and degrees of freedom $2a$
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(c) > 0$ Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item For this distribution, $\chi(\bh)$ shows asymptotic dependence that does not approach 0 as $\bh \rightarrow \infty$ 
   \item This occurs because all observations (near and far) share the same $z_t$ and $\sigma_t^2$.
    \item We deal with this through a daily random partition (similar to Huser and Davison).
  \end{itemize}
\end{frame}

\begin{frame}{Daily random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z_t$ and $\sigma^2_t$ to vary by site.
    \begin{align*}
      Y_t(\bs) = X_t(\bs) \beta + \alpha z_t(\bs) + \sigma(\bs) v_t(\bs)
    \end{align*}
    \item Consider a set of daily knots $w_{tk} \sim$ Uniform that define a random daily partition
    $P_{t1}, \ldots, P_{tK}$ such that
    \begin{align*}
      P_{tk} = \{s : k = \argmin_\ell|| \bs - w_{t\ell}|| \}
    \end{align*}
    \item For $\bs \in P_{tk}$
    \begin{align*}
      z_t(\bs) &= z_{tk}\\
      \sigma^2_t(\bs) &= \sigma^2_{tk}
    \end{align*}
    \item Within each partition $Y_t(\bs)$ has the same MV skew-t distribution as before.
  \end{itemize}
\end{frame}

\begin{frame}{Example daily partition}
	Two sample partitions (number is at partition center)
    \centering
    \includegraphics[width=0.54\linewidth]{./plots/example-partition-1.pdf}
    \includegraphics[width=0.54\linewidth]{./plots/example-partition-2.pdf}
\end{frame}

@@

\begin{frame}{Sample simulated datasets}
  \centering
  \includegraphics[width=1\linewidth]{./plots/gauss-vs-skew-t3.pdf}
\end{frame}

\begin{frame}{Spatiotemporal Model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item We can account for time in one of two ways
    \begin{itemize}
    	\item The mean: e.g. AR(1)
	\item Three dimensional covariance model (e.g. Huser and Davison, 2014) 
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps:
    \begin{enumerate}[1.]
      \item Impute missing observations and censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Model comparisons}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item 9 different analysis methods incorporating
    \begin{itemize}
      \item Gaussian vs $t$ vs skew-$t$ marginal distribution
      \item $K=1$ partition vs $K=5$ partitions
      \item No thresholding vs thresholding at $T=0.90$ sample quantile
    \end{itemize}
    \item All methods use a \Matern or exponential covariance ($\nu = 0.5$)
    \item Compare quantile and Brier scores using 5-fold cross validation (Gneiting and Raftery, 2007)
    \item Mean function modeled using a first-order spatial trend
  \end{itemize}
\end{frame}

\begin{frame}{Quantile score for cross-validation}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item The quantile score for the $\tau$th quantile is
    \begin{align*}
      2 \{ I[y < \widehat{q}(\tau)] - \tau\} (\widehat{q} - y)
    \end{align*}
    where:
    \begin{itemize}
      \item $y$ is a test set value
      \item $\widehat{q}(\tau)$ is the estimated $\tau$th quantile
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Brier score}
  \begin{itemize} \setlength{\itemsep}{0.5em}
	\item The Brier score for predicting exceedance of threshold $c$ is
	\begin{align*}
	  [e(c) - P(c)]^2
	\end{align*}
	where 
	\begin{itemize}
		\item $y$ is a test set value
		\item $e(c) = I[y > c]$
		\item $P(c)$ is the predicted probability of exceeding $c$
	\end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Future work}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Comparison with extreme value analysis methods
    \item Reporting of compliance with EPA ozone standards
    \item Including time in the model via standard spatiotemporal Gaussian models.  
  \end{itemize}
\end{frame}

\begin{frame}{Questions}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Questions?
    \item Thank you for your attention.
    \item Acknowledgment: This work was funded by EPA STAR award R835228.
  \end{itemize}
\end{frame}
